---
title: "5 Data Frames"
author: "Castory Munishi"
date: "10/27/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Frames

+ The most common data structures involved in data analysis.

+ Like lists, data frames can contain multiple data types. 

+ Unlike lists all elements of a data frame are vectors of equal length.

+ The `read_csv()` from readr package is used to import csv data

+ `str()` is used to display the internal structure of an R object

```{r code1, message=FALSE, warning=FALSE, paged.print=TRUE}
library(readr)
recent_grads <- read_csv("data/recent_grads.csv")
#str(recent_grads)
```

## Indexing Data Frames

+ When we want to return a specific column, row, or value from a data frame, we can index them. 
 
+ The syntax for indexing data frames is very similar to that of matrices.
 
+  Let's say we want to return the column of the data frame containing Major.

+  We can perform this operation using several techniques.

## Indexing by Position


```{r code2}
 recent_grads[,3]
```


## Indexing by names

```{r code3}
 recent_grads[,'Major']
```


## Using multiple columns

```{r code4}
recent_grads[,c('Major','Men','Women')]
```

## Indexing using `$` symbol

```{r code5}
recent_grads$Total
```


## Indexing Rows

+ Let's say that instead of a certain column, we want to extract a row of a data frame.

+ To return the fifth row of the data frame, we could write 

```{r code6}
 recent_grads[5,]
```

## Extracting a subset of values

+ As with matrices, we can specify multiple elements to extract from a data frame.
+  Let's extract the first ten rows of the Major and Major_category column 

```{r code7}
recent_grads[c(1:10),c('Major','Major_category')]
```

## Extracting Rows and Saving into a dataframe

+ Extract the first, third, and fifth rows of the Unemployment_rate column and save the output in a variable named unemployment_subset.

```{r code8}
unemployement_subset <- recent_grads$Unemployment_rate[c(1,3,5)]
unemployement_subset
```

## Selecting Columns

+ Our goal is choosing a college major

+ Criteria for choosing a college major
   
   1.You're interested in the physical and life sciences.
   
   2.You'd like a decent salary. Recent graduates must have a median salary above 40,000 USD.
   
   3.You'd like to avoid male-dominated majors. More than 40 percent of graduates must be women.
 
+ We will use the variables in the recent_grads dataframe to make the decesion

## Using `dplyr` package in manipulating Data Frames

+ Use the dplyr function `select()` to create a new data frame with only columns containing the variables we want to keep for our analysis.

+ Using the `select()` function, we'll specify the data frame we want to work with and the columns we want to select

```{r code9, message=FALSE, warning=FALSE}
library(dplyr)
recent_grads_select <- recent_grads %>%
  select(Major, Total, Men, Women, Major_category)
```

## Output

```{r code10}
head(recent_grads_select,5)
```

## The Pipe Operator

+ Notice the `%>%` symbol: This is called the "the pipe operator" 

+ Its purpose is to help you write code that's easier to write, read, and understand. 

+ The pipe lets you write code that chains series of operations together; in the case of the code above

+ We could read the code as, "take the recent_grads data frame and select the Major, Total, Men, Women,  and Major_category columns."

## Selecting the Right Columns

+ Create a new data frame, recent_grads_select, that contains only the following variables from recent_grads:

+  1. Major 2. Major_category 3.Total 4.Men 5.Women 6.Median 7.Unemployment_rate

```{r code11}
recent_grads_select <- recent_grads %>%
  select(Major,Major_category,Men,Women,Total,Median,Unemployment_rate)

#recent_grads_select
```

## Creating New Variables

### The `mutate()` function

+ We can use the mutate function to create a new column based on information contained in existing columns.

+ The mutate function lets us specify the creation of a new column, as well as how the values contained in that new column are calculated:

+  For example adding men percent in the dataset 
```{r code12}
# data_frame_2 <- data_frame_1 %>%
#   mutate(Men_percent = (Men/Total)*100)
```

+ # Within the `mutate()` function, we'll specify a name for our new column and then use `=` to indicate how you want the values that fill it to be calculated.

## Calculating women percent and using `mutate()`


```{r code13}
recent_grads_select <- recent_grads_select %>%
  mutate(Women_percent = round((Women/Total)*100,2))

recent_grads_select$Women_percent
```


## Filtering by Single Criteria

+ Getting closer towards data that meets our criteria

+ Removing data that does not meet our criteria Women_percent >= 40%

+ To remove majors with percentages of women graduates below 40 percent, we'll use another dplyr function: `filter()`

+  We can use `filter()` to specify conditions that values of a variable must meet

## General filter code

```{r code14}
# data_frame_2 <- data_frame_1 %>%
#   filter(Men_percent < 70)

# data_frame_2 <- data_frame_1 %>%
#   filter(Major_category == "Physical Science")
```

## Practice Exercise

+ Filter the recent_grads_select data frame to create a new data frame called recent_grads_money that contains only data on majors where recent graduates' median (the Median variable) salary is greater than 70,000 USD.

## Solution

```{r code15}
recent_grads_money <- recent_grads %>%
  filter(Median > 70000)
recent_grads_money[15:18]
```


## Filtering by Multiple Conditions: Meeting At Least One Criterion

+ What if we wanted to create a new data frame containing college majors that met either of the following criteria:
  
   + Majors in the Engineering catagory `(Major_category == "Engineering")`
   + Majors with over 1,000 total graduates `(Total > 1000)`
   
+ "Majors that are either in the engineering category or have over 1,000 graduates"

+ Archieved by the or operator |


## Example 1

+ Describe what this code does
```{r code16}
recent_grads_engineering1000 <- recent_grads %>%
  filter(Major_category == "Engineering" | Total > 1000)
recent_grads_engineering1000[1:5,c("Major_category","Total")]
```

## Example 2

+ Describe what the code below does

```{r code17}
recent_grads_science <- recent_grads_select %>%
  filter(Major_category == "Engineering" | Major_category == "Physical Science")
recent_grads_science[1:4, c("Major_category","Total")]
```

## Filtering with AND `&` 

+ Both Criteria must be met

+ To use `filter()` to create a new data frame containing only engineering college majors with more than 1,000 total graduates, we'd write:

```{r code18}
recent_grads_engineers1000 <- recent_grads %>%
  filter(Major_category == "Engineering" & Total > 1000)

recent_grads_engineers1000[c("Major_category","Total")]
```

## Winding Up

+ Let's create a new data frame containing only college majors that meet all three of our criteria for college majors.

+ To accomplish this, we'll use `filter()` and the & operator to filter rows in the recent_grads_science data frame to retain only majors that have women as greater than 40 percent of total graduates and median salaries above 40,000 USD.


## Code and Output

```{r code19}
potential_majors <- recent_grads_science %>%
  filter(Women_percent > 40 & Median > 40000)

potential_majors_selected <- recent_grads_science %>%
  filter(Women_percent > 40 & Median > 40000) %>%
  select(Major,Total,Major_category)

print(potential_majors_selected)
```

## Arranging Data Frames by Variables

+ To arrange our data frame rows by TOTAL, we'll use another function in the dplyr package: ``
`arrange()`.

+ The `arrange()` function lets you specify a variable we want to use to re-order the rows of your data frame. Let's say we want to arrange data_frame_1 by Total:

```{r code20}
# data_frame_2 <- data_frame_1 %>% 
#   arrange(Total)
```

+ By default, `arrange()` orders the rows of a data frame by the specified variable in ascending order, from lowest to highest.

+ For descending order use desc()

```{r code21}
# data_frame_2 <- data_frame_1 %>% 
#  arrange(desc(Total))
```

## Using multiple variables to arrange

+  We can arrange data frames by multiple numeric variables as well. If we specify multiple variables, 
+ R will reorder rows of the data frame by each variable in order, moving from left to right.

+ To arrange data_frame_1 in order of ascending Total and descending Men_percent, we could write:

```{r code22}
# data_frame_2 <- data_frame_1 %>% 
#  arrange(Total, desc(Men_percent))
```

## Final Task

+ Create a new data frame, my_majors, containing the data in the potential_majors data frame arranged  in order of increasing Unemployment_rate and decreasing Median (in that order).

```{r code23}
 my_majors <- potential_majors %>% 
   arrange(Unemployment_rate, desc(Median))
my_majors[,1:6]
```