---
title: "Control Structures"
author: "Castory Munishi"
date: 10/27/2020
execute: 
  echo: true
  warning: false
format: revealjs
editor: visual
---


```{r setup, include=FALSE}
pacman::p_load(kableExtra)
```

## Introduction to Control Structures

### Introduction to the data

```{r code1, message=FALSE, warning=FALSE}
pacman::p_load(readr)
scores <- read_csv("datasets/scores.csv")
kable(scores[1:4,c("home_country","home_goals","away_country",
             "away_goals","win_country")])
```

+ Our goal is to add a new variable to the data frame that provides information about whether the home team won each match.

## Steps in The analysis

+ We will start by working with just two rows of the dataframe as practice and we will follow these steps:

 1. Analyze the data to determine whether the home team won each match.
 2. Create a vector containing either "home team won" or "home team did not win" for each row  based on the analysis.
 3. Add this vector to the data frame as a new column
 
## Which team won?

+ Subset the first two rows of the df to a tibble named scores_two

```{r code2}
scores_two <- scores[1:2,]
```

+ In the first row, the value in the home_goals column equals 3 and the value in away_goals column equals 1,  so we can determine that the home team won and the same can be done for the second row.

+ Create a string vector containing home team won and add it to scores_two, then subset the first five rows of scores

## Code

```{r code3}
library(dplyr)
home_team_won <- c("Home Team Won","Home Team Won")
scores_two <- scores_two %>%
  mutate(home_team_won = home_team_won )
scores_five <- scores[1:5,c("home_country","home_goals","away_country",
                            "away_goals","win_country")]
kable(scores_five)
```

## Another Vector

+ Create a new vector called home_team_won that contains either "home team won" or "home team did not win" for each match in scores_five (in order).

```{r code4}
home_team_won <- c("home team won","home team won",
                   "home team did not win",
                   "home team won","home team won")
```


## Adding the new Vector

+ Use the `mutate()` function to add the home_team_won vector to the scores_five data frame. Name the column home_team_won.

```{r code5}
scores_five <- scores_five %>%
  mutate(home_team_won = home_team_won)
```


## Selection : Writing Conditional Statements

+ Will return a given value if the condition is satisfied

+ For our case looking at whether the home team won or not is when home goals > away goals is true.

+ This is done by using **if statement**

+ Let's write an if statement to return "home team won" if home_goals > away_goals during the first match.

```{r code6}
if (scores$home_goals[1]>scores$away_goals[1]){
   print("home team won")
}
```


## if else statement

+ What about when the home team loses, though?

+ In the example given,  if the away team scored more goals than the home team, nothing would be returned.

+ To return "home team did not win" if the home team did not score more goals than the away team,  you can add a type of conditional statement called an else statement to your code.

```{r code7}
if (scores$home_goals[1]>scores$away_goals[1]){
   print("home team won")
}else{
   print("home team did not win")
}
```

## Task

+  Write an if-else statement to return:
 
  1."home team won" if the number of goals scored by the home team is greater   than the number of goals scored by the away team
  2. "home team did not win" if the condition specified above is not met
  
## Solution

```{r code8}
if(scores$home_goals[3]>scores$away_goals[3]){
   print("home team won")
}else{
   print('home team did not win')
}
```


## Repetition : For-Loops

+  The previous method of writing code is repititive

+  Generally, in programming, **it's best to minimize repetition in your code as much as possible**. 

+  If you find yourself copying and pasting blocks of code several times, it's time to seek a better solution.

           **Use for loops**

## Example

+ Let's write a for-loop to print every number in a sequence of numbers from one to 5.

```{r code9}
for( i in 1:5){
  print(i)
}
```

+ The index variable i represents an element of a sequence. You can read the code above as, 
 "For every element in the sequence of the numbers one to 5, print the element".
 
## Using Variable Names

+ You can use any variable name you want instead of i for the index

+ Should consider using a name that describes what the variable actually represents to make complex code more readable.

+ Next, let's write a for-loop to print the date of the first five matches in in the World Cup scores data frame:

```{r code10}
for( d in scores$match_date[1:5]){
    print(d)
}
```


## Task 

+ Write a for-loop to print each element in the column home_country of the scores data frame.

## Solution

```{r code11}
# for ( h in scores$home_country){
#  print(h)
# }
```

## Looping Over raws of a data frame

+ When you write a for-loop, the elements you specify can be values, vectors, lists, or other data structures.

+ Since you are working with a data frame (scores), let's write a for-loop to execute an operation on elements that are rows of the data frame

+ When you write a for-loop, the elements you specify can be values, vectors, lists, or other data structures.

## Our For Loop

+ The for-loop will calculate the total number of goals (away_goals + home_goals) for each match

+ In the scores data frame, each match has its own row. Since you want to perform the addition operation for each row of the data frame,

+  the first part of the for-loop will consist of defining i as an element of the sequence of numbers from one to the number of rows in  the data frame. 

+ Here's how you'll specify the sequence: `for ( i in 1:nrow(scores))`

+ "for each element in the sequence one to 64"

+ In programming, it's good practice to refer to data objects instead of including numbers in your code. 

## The Loop

```{r code12, echo=TRUE}
# for ( i in 1:nrow(scores)){
#  print(scores$home_goals[i] + scores$away_goals[i])
#}
```

## Task 

Write a for-loop to subtract away_goals from home_goals for each match in scores. 

## Solution

```{r code13}
# for (i in 1:nrow(scores)){
#  print(scores$home_goals[i] - scores$away_goals[i])
# }
```


## Nested Control Structures 1/2

+ Control structures are powerful tools, and can do even more when combined together. 

+  For example, the actions repeated by a for-loop don't need to be sequential code — they could be conditional statements.
+ This pattern of executing one or more control structures inside another one is called **nesting**.

+ Objective : Originally, we wanted to create a new variable that provides information about whether the home team won or lost each match.


## Code

```{r code14}
for (i in 1:nrow(scores)){
  if(scores$home_goals[i] > scores$away_goals[i]){
    print("home team won")
  }else{
    print('home team did not win')
  }
}
```


## Nested Control Structures 2/2

+ Nested control structures are very powerful, but can be tricky to get used to and can make finding errors in your code harder. 

+  The workflow in the previous examples is a good way to avoid problems when writing nested control structures: 

+ Start by writing the inner operation and make sure it works correctly, and then wrap the outer operation around it.

+ Next, you'll learn to store the output of a loop in a variable. Let's write a for-loop that outputs logical data, TRUE and FALSE, that would make a useful variable.


## Task

+ Write a for-loop that, for each row in the scores data frame, prints TRUE if home_goals is greater than away_goals and FALSE if not

```{r code15}
for ( i in 1:nrow(scores)){
  if(scores$home_goals[i] > scores$away_goals[i]){
    print(TRUE)
  }else{
    print(FALSE)
  }
}
```


## Storing For-loop Output in Objects

+ We have just been printing the output of a for loop you'll want to use the output of a for-loop instead of just printing it.

+ calculating total scores

```{r code16}
for ( i in 1:nrow(scores)){
   print(scores$home_goals[i] + scores$away_goals[i])
}
```

## You need a empty vector

+ To modify the for-loop to store the total goals scored in each match in a vector instead of printing it, you'll need to first create an empty vector to add values to:

+ Then, you'll write the for-loop to include the following code to add new elements to the vector:

```{r code17}
total_goals <- c()
# total_goals <- c(total_goals,"new element")
```

## New elements

+ The new elements you'll add to the vector are the sums of home_goals and total_goals for each match.
 
```{r code}
for ( i in 1:nrow(scores)){
  total_goals <- c(total_goals,scores$home_goals[i] + scores$away_goals[i])
}
total_goals[1:5]
length(total_goals)
```


## Winning
Let's create a new, empty vector, home_team_won, to store the logical output of the for-loop

`r home_team_won <- c()` #single line code

+ In place of the print() functions in the for-loop, you can include code to store output in home_team_won:

 `home_team_won <- c(home_team_won, TRUE)`
 `home_team_won <- c(home_team_won, FALSE)`
 

## Task 

+ Write a for-loop that, for each row in the scores data frame,  outputs TRUE if home_goals is greater than away_goals and FALSE if not.

+ Store the output in a vector called home_team_won.

## Solution

```{r code19}
for ( i in 1:nrow(scores)){
  if(scores$home_goals[i] > scores$away_goals[i]){
    home_team_won <- c(home_team_won,TRUE)
  }else{
    home_team_won <- c(home_team_won,FALSE)
  }
}
home_team_won[1:5]
```

## Adding to the dataset

+ You've now created a new variable, home_team_won, that provides information about whether the home team won each match in the data set.

+ You can use this new information to begin exploring whether teams playing in the 2014 World Cup experienced a home advanage.

```{r code20}
scores <- scores %>%
  mutate (home_team_won = home_team_won)
```

## Handling more than two cases

+ It would be more informative to have a variable containing three outcomes for the home team:

    1. Won   home_goals > away_goals
    
    2. Lost  home_goals < away_goals
    
    3. Tied  home_goals = away_goals

+  Instead of writing code to specify two possible outcomes, in this case, you'll need to specify three. __else if__

## Else if

```{r code21}
if (scores$home_goals[1] > scores$away_goals[1]){
  print("win")
}else if(scores$home_goals[1] < scores$away_goals[1]){
  print('lose')
}else if(scores$home_goals[1] < scores$away_goals[1]){
  print("tied")
}
```

## Default Action

+ You can chain together as many else if statements in your code as needed.

+ In the scores data set, there are only three possible match outcomes.
#
+ However, if you needed to specify a default action to perform if none of the prior conditions are met,you can include an else statement in your code.


```{r code22}
if (scores$home_goals[1] > scores$away_goals[1]){
  print("win")
}else if(scores$home_goals[1] < scores$away_goals[1]){
  print('lose')
}else if(scores$home_goals[1] < scores$away_goals[1]){
  print("tied")
}else{
  print("no action")
}
```


## Example : Third Match

+ What does this code do?

```{r code23}
if (scores$home_goals[3] > scores$away_goals[3]){
  print("win")
}else if(scores$home_goals[3] < scores$away_goals[3]){
  print('lose')
}else if(scores$home_goals[3] == scores$away_goals[3]){
  print("tied")
}
```

## More than Two Cases : Writing a for loop

1. Create a new vector, home_team_result, that contains "win", "lose", or "tie" for the home team for each match in scores.

2. Overwrite the scores data frame with a new scores data frame containing the home_team_result variable. 


## The code

```{r code24}
home_team_result <- c()
for(i in 1:nrow(scores)){
  if(scores$home_goals[i] > scores$away_goals[i]){
    home_team_result <- c(home_team_result,"win")
  }else if(scores$home_goals[i] < scores$away_goals[i]){
   home_team_result <- c(home_team_result,"lose")
  }else if(scores$home_goals[i] == scores$away_goals[i]){
   home_team_result <- c(home_team_result,"tied")
  }
}
scores <- scores %>%
  mutate( home_team_result = home_team_result)
```


## Remarks 

+ If-else statements and for-loops are programming tools common to practically all programming languages

+ It's important to understand how they work and when to use them.

+ R is characterized by some special features superior to for-loops  when it comes to repeating the same operation on multiple elements