---
title: "Stratified and Cluster Sampling"
author: "Castory Munishi"
date: 11/05/2022
execute: 
  echo: true
  warning: false
format: revealjs
editor: visual
---


## Challenge of Simple Random Sampling

+ SRS will not always result in a sample having all groups since its completely random
+ Sometimes we might want to analyze patterns for each possible group
+ In our example dataset we have five different positions.
+ F: Foward, G: Guard, C: Center, G/F: Guard/Foward, F/C: Foward/Center

## SRS Missing a category

+ Using SRS we might result in a situation like

![](images/s1m1_srs_down.svg)

## Getting Group Representation

+ Group representation can be obtained by *stratified sampling*
+ The dataset is organized into different groups
+ The randomly sample from each group
+ Each individual stratified group known as a *stratum*
+ Multiple groups known as *strata*


![](images/s1m1_stratified_sampling_v3.svg)


## `sample_n()` function

+ Enables sampling rows.
+ Takes two arguments, the dataset and number of rows to be sampled

```{r}
pacman::p_load(
  readr,
  dplyr,
  ggplot2
)

ds <- read_csv("datasets/wnba.csv")

# getting ten rows
ds_sampled <- sample_n(ds,size=10)
```


## Task 1

Calculate the average age and the average number of games played from a random sample. 

i. Designate set.seed(1) to make the results reproducible.
ii. Sample 30 rows from the wnba dataframe. Save the results as thirty_samples.
iii. Calculate the average age of this sample group. Assign the results to mean_age.
iv. Calculate the average number of games played for this sample group. Assign the results to mean_games.


## Solution 1

```{r}
set.seed(1)
ds_30 <- sample_n(ds,size=30)
mean_age <- mean(ds_30$Age, na.rm=T)
mean_games <- mean(ds_30$Games_Played, na.rm = T)
```

+ The mean age is `r mean_age`
+ The mean number of games `r mean_games`


## Creating and Analyzing Strata with dplyr

+ When can use the `group_by()` function to create strata
+ We can then apply one or more functions for each strata


## Getting Mean by Stratum

```{r}
set.seed(1)
ds %>% 
  # Split: stratify by player position
  group_by(Pos) %>% 
  # Apply: sample 10 observations for each player position stratum
  sample_n(10) %>%
  # Apply & combine: calculate average points scored for each stratum, combine results
  summarize(mean_pts = mean(PTS))
```

