---
title: "US Health Disparities by Race"
subtitle: "Proposed Solution"
date: 03/09/2023
author: "Kili Analytics"
format:
  html:
    theme: united
    code-fold: true
    code-tools: true
editor: visual
toc: true
---

## 1. Data Cleaning and Preparation

### Loading Packages and Data

-   Importing the dataset

```{r}
pacman::p_load(
  rio, # for data import
  here, # for file paths
  janitor, # for cleaning and simple tables
  magrittr, # for pipes
  forcats, # for working with factors
  epitools, # for odds ratios
  gtsummary, # for summary tables
  epikit, # for age categories
  flextable, # for making tables
  tidyverse # for data manipulation and plotting
)

nhpi_raw <- import("nhpi2014.csv") %>%
  select(ID, everything()) # making id the first column
```

### Cleaning functions

-   The code below contains custom functions for data cleaning

```{r}
# custom recoding functions

recoding <- function(x){
  ifelse( x == 1,4,
          ifelse(x == 2,3,
                  ifelse(x == 3,2,
                         ifelse( x == 4,1,
                                 ifelse(
                                   x == 5,0,NA)))))
}

# columns to exclude if they have NAS
binary_vars <- c("CURRSMOK", "ACTIVE", "HTN", "DIAB", "ASTHMA", "JOINTPAIN", "FEMALE_SEX", "EMPLOYED")

exclude_cols = c("DISTRESS_SCORE", "MARRIED", "BMI_CAT", "ALCOHOL", binary_vars )


# a function for recoding binary variables
binary_recode <- function(x){
  if_else(x == 1, "Yes", "No")
}

# a function for recoding Kessler Screening Scale for Pyschological Distress questions (k-6)
kessler_vars <- c("ASISAD", "ASINERV", "ASIRSTLS",
                  "ASIHOPLS", "ASIEFFRT", "ASIWTHLS")

kessler <- function(x){
  ifelse( x == 1,"All the time",
          ifelse(x == 2,"Most of the time",
                  ifelse(x == 3,"Some of the time",
                         ifelse( x == 4,"A little of the time",
                                 ifelse(
                                   x == 5,"None of the time",NA)))))
}

## a function for custom fisher test
fisher.test.simulate.p.values <- function(data, variable, by, ...) {
  result <- list()
  test_results <- stats::fisher.test(data[[variable]], data[[by]], simulate.p.value = TRUE)
  result$p <- test_results$p.value
  result$test <- test_results$method
  result
}
```

### The cleaning pipe chain

The code below contains a pipe chain of data cleaning functions according to the instructions of the exercise.

-   Note: `across()` function is a powerful `dplyr` function for changing multiple columns at once

-   On multiple steps in the code below new variables are defined by using the `mutate()` function

```{r}
nhpi <- nhpi_raw %>%
  # recoding 7,8, 9 and 9999 into NAs
   mutate(across(.cols = everything(),
                ~ ifelse(.x %in% c(7:9,9999), NA, .x)
                )) %>%
  # creating age categories (epikit package)
  mutate(AGE_CAT = age_categories(AGE_P, 
                                  breakers = c(18,45,64))) %>%
  # creating BMI categories
  mutate(BMI_CAT = case_when(
    BMI < 18.5               ~ "Underweight",
    BMI >= 18.5 & BMI < 25   ~ "Normal Weight",
    BMI > 25 & BMI < 30      ~ "Overweight",
    BMI >= 30                ~ "Obese",
    is.na(BMI)               ~ "Unknown BMI"
  ),
  BMI_CAT = fct_relevel(
    BMI_CAT, "Underweight", "Normal Weight",
    "Overweight", "Obese", "Unknown BMI"
  )) %>%
  # Recoding Maritial Status
  mutate(MARRIED = case_when(
    MAR_STAT == 4 ~ "Never Married",
    MAR_STAT == 2 | MAR_STAT ==3 ~ "Widowed, divorced, or separated",
    MAR_STAT == 1 | MAR_STAT == 5 ~ "Married or living with partner"
  )) %>%
  # recoding alcohol use variable
  mutate(ALCOHOL = case_when(
    ALCOHOL == 0 ~ "None",
    ALCOHOL == 1 ~ "Light Drinker",
    ALCOHOL == 2 ~ "Moderate Drinker",
    ALCOHOL == 3 ~ "Heavy Drinker"
  ),
  ALCOHOL = fct_relevel(
    ALCOHOL, "None", "Light Drinker",
    "Moderate Drinker", "Heavy Drinker"
  )) %>%
  # recoding distress variables
  mutate(
    ASISAD2     = recoding(ASISAD),
    ASINERV2    = recoding(ASINERV),
    ASIRSTLS2   = recoding(ASIRSTLS),
    ASIHOPLS2   = recoding(ASIHOPLS),
    ASIEFFRT2   = recoding(ASIEFFRT),
    ASIWTHLS2   = recoding(ASIWTHLS)
  ) %>%
  # Calculating Participant distress score
  mutate(
    DISTRESS_SCORE = ASISAD2 + ASINERV2 + ASIRSTLS2 + ASIHOPLS2 + ASIEFFRT2 + ASIWTHLS2
  ) %>%
  # Creating outcome variables
  mutate(
    DISTRESS_SEVERE = case_when(
      DISTRESS_SCORE >= 13 ~ "Severe Pyschological Distress",
      DISTRESS_SCORE < 13  ~ "No Severe Pyschological Distress"
    ),
    # Converting DISTRESS_SEVERE to factor variable
    DISTRESS_SEVERE = fct_relevel(
      DISTRESS_SEVERE, "No Severe Pyschological Distress",
                       "Severe Pyschological Distress"
    )) %>%
  mutate(
    # converting DISTRESS_LEVEL to factor variable
    DISTRESS_LEVEL = case_when(
      DISTRESS_SCORE == 0   ~ "No Distress",
      DISTRESS_SCORE <= 5  ~ "Low Distress",
      DISTRESS_SCORE <= 10 ~ "Moderate Distress",
      DISTRESS_SCORE <= 24 ~ "High Distress"
    ),
    DISTRESS_LEVEL = fct_relevel(
      DISTRESS_LEVEL, "No Distress", "Low Distress", 
                     "Moderate Distress", "High Distress"
    )) %>%
  # recoding binary variables
  mutate(across(.cols = all_of(binary_vars),
                ~ binary_recode(.x))) %>%
  mutate(across(.cols = all_of(kessler_vars),
                ~ kessler(.x))) %>%
  # removing NAs in selected columns
  drop_na(all_of(exclude_cols)) 

# Cleaning testing code

# nhpi %>%
#  tabyl(DISTRESS_LEVEL)
  
```

## Statistical Analysis

### Descriptive Statistics

The table below contains descriptive characteristics of the study participants

```{r}
#| warning: false
#| message: false
nhpi %>%
  select(AGE_CAT, FEMALE_SEX, EMPLOYED, MARRIED, CURRSMOK, ACTIVE, ALCOHOL, BMI_CAT, DIAB, HTN, JOINTPAIN, ASTHMA, DISTRESS_LEVEL, AGE_P) %>%
  tbl_summary(
    by = DISTRESS_LEVEL,
    label = list(
      AGE_CAT    ~ "Age Category",
      FEMALE_SEX ~ "Female",
      EMPLOYED   ~  "Currently Employed",
      MARRIED    ~  "Marital Status",
      CURRSMOK   ~  "Current Smoker (Yes)",
      ACTIVE     ~  "Moderate or High Physical Activity Level (Yes)",
      ALCOHOL    ~ "Alcohol Use",
      BMI_CAT    ~ "BMI Category",
      DIAB       ~ "Diabetes (Yes)",
      HTN        ~ "Hypertension (Yes)",
      JOINTPAIN  ~ "Jointpain (Yes)",
      ASTHMA     ~ "Asthma (Yes)"
    ),
    statistic = list(all_continuous() ~ "{mean} ({sd})"),
    digits = everything() ~ c(0,1)
  ) %>%
  add_n() %>%
  add_p(test = BMI_CAT ~ "fisher.test.simulate.p.values") %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  modify_caption("**Table 1: Participant Summary Characteristics**") %>%
  as_flex_table()

```

1.	Based on your results in Table 1, which risk factors showed a statistically significant association with psychological distress level?
The factors that showed statistically significant results are:
 
a)	Age Group
b)	Female Sex
c)	Employment Status
d)	Marital Status
e)	Smoking Status
f)	Physical Activity level
g)	Alcohol Use
h)  BMI Category
h)	Diabetes
i)	Hypertension
j)	Joint Pain
k)	Asthma

2. You are interested in further exploring the association between marital status and psychological distress level.



a. Is there an association between marital status (MARRIED) and psychological distress level (DISTRESS_LEVEL)? Report the name of the test, the null and alternative hypotheses, test statistic, degrees of freedom, p-value, and your conclusion.

```{r}
nhpi %>%
  tabyl(MARRIED, DISTRESS_LEVEL) %>%
  adorn_totals(where = "row") %>%
  adorn_percentages() %>%
  adorn_pct_formatting() %>%
  adorn_ns(position = 'front') %>% # display as: count (percent)
  qflextable()
```

```{r}
nhpi %>%
  tabyl(MARRIED, DISTRESS_LEVEL) %>%
  chisq.test()
```

i.	There is an association between marital status and psychological distress level

ii.	The test used is Chi Square test as these are categorical variables

iii. Hypothesis

- The Null Hypothesis: There is no association between marital status and psychological distress level

- The Alternative Hypothesis: There is no association between marital status and psychological distress level

iv.	Test Statistic: `38.692`

v.	Degrees of freedom: `6`

vi.	P-value: `< 0.001`

> vii.	People who are never married have more psychological distress as compared to married people


b. Is there an association between marital status (MARRIED) and severe psychological distress (DISTRESS_SEVERE)? Report the name of the test, the null and alternative hypotheses, test statistic, degrees of freedom, p-value, and your conclusion. Calculate and interpret the appropriate measures of effect (+95% confidence intervals), using “married” as the reference group.

```{r}
nhpi %>%
  tabyl(MARRIED, DISTRESS_SEVERE) %>%
  adorn_totals(where = "row") %>%
  adorn_percentages() %>%
  adorn_pct_formatting() %>%
  adorn_ns(position = 'front') %>% # display as: count (percent)
  qflextable()
```
```{r}
nhpi %>%
  tabyl(MARRIED, DISTRESS_SEVERE) %>%
  chisq.test()
```

```{r}
oddsratio(nhpi$MARRIED, nhpi$DISTRESS_SEVERE, 
          method = "wald", conf.level = 0.95)$measure

```

i.	There is an association between marital status and severe psychological distress

ii.	The test used is Chi square test as these are categorical variables

iii. Hypothesis

- The Null Hypothesis: There is no association between marital status and severe psychological distress

- The Alternative Hypothesis: There is association between marital status and severe psychological distress

iv.	Test Statistic: `8.5765`

v.	Degrees of freedom: `2`

vi.	P-value:  `0.01373`

vii.	People who are never married are more likely to have severe psychological distress as compared to married people

> viii.	Odds Ratio (95% CI)

> Never Married `1.85 (1.17 – 2.92)`

> Married/ Living with Partner `1.67 (0.99 – 2.8)`


### Inferential Statistics

Generate a multiple linear regression model to determine which risk factors are associate with the psychological distress score

```{r}
model1 <- lm(DISTRESS_SCORE ~ AGE_P + FEMALE_SEX + MARRIED +
               BMI + CURRSMOK + ACTIVE + ALCOHOL,
             data = nhpi)

model1 %>%
  tbl_regression(
    label = list(
      AGE_P      ~ "Participant's Age",
      FEMALE_SEX ~ "Sex (Female)",
      MARRIED    ~ "Marital Status",
      CURRSMOK   ~ "Current Smocker",
      ACTIVE     ~ "Physically Active",
      ALCOHOL   ~ "Uses Alcohol"
    )
  ) %>%
  bold_p() %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  modify_caption("**Table 2: Risk factors of Pyschological Distress Score**") %>%
  as_flex_table()

```

Based on your results in Table 2, which risk factors showed a statistically significant association with psychological distress score?

The Risk factors that showed a statistically significant association with psychological distress score are :

a)	Age

b)  Female Sex

c)	Marital Status

d)  BMI

e)	Physical Activity

f)	Alcohol Use


```{r}
model2 <- glm(DISTRESS_SEVERE ~ AGE_P + FEMALE_SEX + MARRIED + 
               BMI + CURRSMOK + ACTIVE + ALCOHOL,
              family = binomial(link = logit),data = nhpi)

model2 %>%
  tbl_regression(
    exponentiate = TRUE
  ) %>%
  bold_p() %>%
  modify_header(label = "**Variable**") %>%
  modify_caption("**Table 3: Risk factors of Severe Pyschological Distress**") %>%
  as_flex_table()
```

Based on your results in Table 3, which risk factors showed a statistically significant association with severe psychological distress?

The factors which have shown statistically significant association with severe psychological distress are:

a) Smocking Status

b) Physical Activity